---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

<!--
source of this file: https://github.com/nuest/RepData_PeerAssessment1

build with knitr::knit2html() does not work, use rmarkdown::render("PA1_template.Rmd")

working directory is assumed to be the one with this file
list.files(getwd())
-->

## Loading and preprocessing the data
```{r, echo=TRUE}
datafile <- "./activity.csv"
if(!file.exists(datafile)) {
  utils::unzip("./activity.zip")
}
activity_rawdata <- read.csv(file = datafile)
str(activity_rawdata)
head(activity_rawdata)
tail(activity_rawdata[1000:1100,])
```

## What is the mean total number of steps taken per day?

### 1. Total number of steps per day

```{r, echo=TRUE}
library(dplyr, quietly = TRUE, warn.conflicts = FALSE)
activity_stepsperday <- activity_rawdata %>% group_by(date) %>% summarise(steps = sum(steps))
head(activity_stepsperday)
```

### 2. Histogram

```{r, echo=TRUE}
hist(activity_stepsperday$steps, breaks = 24)
```

### 3. Mean and median

```{r, echo=TRUE}
mean_steps_per_day <- mean(activity_stepsperday$steps, na.rm = TRUE)
mean_steps_per_day
median_steps_per_day <- median(activity_stepsperday$steps, na.rm = TRUE)
median_steps_per_day
```

The rounded mean number of steps per day is `r toString(round(mean_steps_per_day))`, the median is `r median_steps_per_day`.

## What is the average daily activity pattern?

### 1. Time series plot

```{r, echo=TRUE}
activity_stepsperinterval <- activity_rawdata %>% group_by(interval) %>% summarise(steps = mean(steps, na.rm = TRUE))
summary(activity_stepsperinterval)
head(activity_stepsperinterval)
plot(x = activity_stepsperinterval$interval, y = activity_stepsperinterval$steps, type = "l",
     xlab = "Minute in the day", ylab = "Average number of steps", main = "Average number of steps",
     sub = paste("[Averaged over", dim(activity_stepsperday)[1], "days]"))

top_intervals <- activity_stepsperinterval %>% top_n(steps, n = 3)
```

### 2. Top interval

The 5-min interval with the maximum number of steps (on average across all days) is the interval starting at minute `r top_intervals[[1,1]]`.

## Imputing missing values

### 1. Number of missing values

```{r, echo=TRUE}
sum(is.na(activity_rawdata$steps))
```

### 2. Strategy for missing values

```{r, echo=TRUE}
replace_na_with_interval_mean <- function(x) {
  replacement_value <- x[["steps"]]
  if(is.na(x[["steps"]])) {
    # fill in average for that interval
    #cat(as.numeric(x[["interval"]]), "\n")
    replacement_value <- activity_stepsperinterval[activity_stepsperinterval$interval==as.numeric(x[["interval"]]),2]$steps
    #cat(x, "using replacement", replacement_value, "\n")
    #steps <- as.numeric(replacement_value)
  }
  as.numeric(replacement_value)
}
```

### 3. Filled data set

```{r, echo=TRUE}
activity_filledsteps <- apply(X = activity_rawdata, MARGIN = 1, FUN = replace_na_with_interval_mean)
# replace steps values
activity_filled <- activity_rawdata
activity_filled[,c("steps")] <- activity_filledsteps

# no NAs?
sum(is.na(activity_filled$steps))
summary(activity_filledsteps)
```

### 4. Analysis of filled dataset

```{r, echo=TRUE}
# create histrogram
activity_filled_stepsperday <- activity_filled %>% group_by(date) %>% summarise(steps = sum(steps))
hist(activity_filled_stepsperday$steps, breaks = 24)
```

Mean and median of the filled dataset:

```{r, echo=TRUE}
mean(activity_filled_stepsperday$steps)
median(activity_filled_stepsperday$steps)
```

The values do not differ considerably, the mean stayed the same and the median adjusted to the mean.

The total number of steps changed from `r format(sum(activity_rawdata$steps, na.rm = TRUE))` to `r format(sum(activity_filled$steps))` (factor of `r sum(activity_filled$steps)/sum(activity_rawdata$steps, na.rm = TRUE)`).

<!--
```{r}
plot(x = activity_stepsperday$steps, y = activity_filled_stepsperday$steps, 
     title = "Steps per day vs. filled steps per day")
```
-->

## Are there differences in activity patterns between weekdays and weekends?

```{r}
# replace day name with our own category
replace_day_with_type <- function(day) {
  #cat(day, day %in% c("Saturday", "Sunday"), "\n")
  if(day %in% c("Saturday", "Sunday")) {
    return("weekend")
  }
  else {
    return("weekday")
  }
}

# set the day
activity_pattern <- cbind(activity_filled, day = 
                            sapply(X = weekdays(as.Date(activity_filled$date)),
                                   FUN = replace_day_with_type))
str(activity_pattern)
average_weekdays <- mean(activity_pattern[activity_pattern$day == "weekday",c("steps")], na.rm = TRUE)
average_weekend <- mean(activity_pattern[activity_pattern$day == "weekend",c("steps")], na.rm = TRUE)

#activity_daytype <- activity_pattern %>% group_by(day) %>% summarise(steps = mean(steps, na.rm = TRUE))
#str(activity_daytype)
plot(x = activity_pattern[activity_pattern$day == "weekday",c("interval")],
     y = activity_pattern[activity_pattern$day == "weekday",c("steps")],
     type = "l",
     xlab = "Minute in the day", ylab = "Number of steps", main = "Number of steps on weekdays")
plot(x = activity_pattern[activity_pattern$day == "weekend",c("interval")],
     y = activity_pattern[activity_pattern$day == "weekend",c("steps")],
     type = "l",
     xlab = "Minute in the day", ylab = "Number of steps", main = "Number of steps on weekends")
```
